New cardinality estimation algorithms for HyperLogLog sketches

### 2016-05-26 paper/paper.tex
fixed issue in corrected raw estimate + other changes
benchmark chart for corrected raw estimation algorithm
* completed new corrected raw estimation algorithm * added corresponding benchmark chart * modified paper title

\title{Cardinality estimation methods for HyperLogLog sketches}
\subtitle{Draft version}
\author{Otmar Ertl}

@@

\subsection{Correcting the \emph{raw} estimate}
Given a HyperLogLog sketch with its multiplicity vector $\vec{\symCount}$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCount'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and use them instead in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}. For $\symRegVal\in[1,\symRegRange]$ where $\symCount_\symRegVal = \symCount'_\symRegVal$ we have the trivial estimators 
\begin{equation}
\hat{\symCount}'_\symRegVal := \symCount_\symRegVal\quad \text{for $\symRegVal \in [1, \symRegRange]$}.
\end{equation}
To get estimators for all other $\symRegVal$, we consider their expectations
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}\right).
\end{equation}
From \eqref{equ:register_value_distribution} we know that $\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)=
e^{-\frac{\symPoissonRate}{\symNumReg}}$ and $\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)=
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}$, and therefore, we can also write
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg
\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}
\left(1-\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
\end{equation}
and
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg
\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}
\left(1-\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}\right),
\end{equation}
which motivates us to use
\begin{equation}
\hat{\symCount}'_\symRegVal
:=
\symNumReg
\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}
\left(1-\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
\end{equation}
as estimator for $\symRegVal \leq 0$ and
\begin{equation}
\hat{\symCount}'_\symRegVal
:=
\symNumReg
\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}
\left(1-\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}\right)
\end{equation}
as estimator for $\symRegVal \geq \symRegRange+1$, respectively.

Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} gives
\begin{equation}
\label{equ:correctedestimator}
\symPoissonRateEstimate 
= 
\frac{\symNumReg^2}{2\log(2)\sum_{\symRegVal=-\infty}^\infty \hat{\symCount}'_\symRegVal 2^{-\symRegVal}}
=
\frac{\symNumReg^2}
{2\log(2)
\left(
\symSmallCorrectionFunc(\symCount_0) + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symLargeCorrectionFunc(\symCount_{\symRegRange+1}) 2^{-(\symRegRange+1)}
\right)}
\end{equation}
where $\symSmallCorrectionFunc(\symCount_0)$ and $\symLargeCorrectionFunc(\symCount_{\symRegRange+1})$ replace $\symCount_0$ and $\symCount_{\symRegRange+1}$ in the \emph{raw} estimate \eqref{equ:intermediate_estimate}, respectively. The functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are given by
\begin{equation}
\symSmallCorrectionFunc(\symCount) := \symNumReg
\cdot
\left(
\frac{\symCount}{\symNumReg}
+
\frac{1}{2}\sum_{\symRegVal=1}^\infty
\left(\frac{\symCount}{\symNumReg}\right)^{2^\symRegVal} 2^{\symRegVal}
\right)
\end{equation}
and
\begin{equation}
\symLargeCorrectionFunc(\symCount)
:=
\symNumReg
\cdot
\left(
\sum_{\symRegVal=1}^\infty
\left(1-\frac{\symCount}{\symNumReg}\right)	^{2^{-\symRegVal}}
2^{-\symRegVal}
-
\left(
1-\frac{\symCount}{\symNumReg}
\right)
\right)
=
\symNumReg
\cdot
\left(
\frac{\symCount}{\symNumReg}
-
\sum_{\symRegVal=1}^\infty
\left(1-\left(1-\frac{\symCount}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
2^{-\symRegVal}
\right)
\end{equation}

For the special case of linear probabilistic counting with $\symRegRange=0$, we get for the  estimator \eqref{equ:correctedestimator}
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symNumReg}{\log(2)\sum_{\symRegVal=-\infty}^\infty
\left(\frac{\symCount_0}{\symNumReg}\right)^{2^\symRegVal}2^\symRegVal}
=
\frac{\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right)}{\symPowerSeriesFunc\left(\log_2\left(\log\left(\frac{\symNumReg}{\symCount_0}\right)\right)\right)}
\approx
\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right).
\end{equation}
Here we used function $\symPowerSeriesFunc$ defined in \eqref{equ:power_series_function} which is known to be approximately equal to 1. Hence, for this special case we get as expected almost the same estimator as given in \eqref{equ:small_range_estimate}.
\subsection{Estimation algorithm}
\begin{algorithm}
\caption{Cardinality estimation algorithm based on the corrected \emph{raw} estimate.}
\label{alg:estimate_raw_corrected}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symCount}$}
\State $\symNumReg \gets \left\Vert\vec{\symCount}\right\Vert_1$
\State $\symZ\gets 0.5\cdot \symLargeCorrectionFunc(\symCount_{\symRegRange+1})$
\For{$\symRegVal\gets \symRegRange, 1$}
\State $\symZ\gets\symZ + 0.5\cdot \symCount_\symRegVal$
\EndFor
\State $\symZ\gets\symZ+\symSmallCorrectionFunc(\symCount_0)$
\State \Return$\symNumReg^2 / \left(2\log(2)\symZ\right)$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{avg_exec_time}
\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
\label{fig:avg_exec_time}
\end{figure}

>>>>

\title{New cardinality estimation algorithms for HyperLogLog sketches}
\subtitle{Draft version}
\author{Otmar Ertl}

@@

\subsection{Correcting the \emph{raw} estimate}
Given a HyperLogLog sketch with its multiplicity vector $\vec{\symCount}$, we try to find estimates $\hat{\symCount}'_\symRegVal$ for $\symCount'_\symRegVal$ for all $\symRegVal\in\mathbb{Z}$ and use them instead in the estimation formula \eqref{equ:raw_estimate_with_multiplicities}. For $\symRegVal\in[1,\symRegRange]$ where $\symCount_\symRegVal = \symCount'_\symRegVal$ we have the trivial estimators 
\begin{equation}
\hat{\symCount}'_\symRegVal := \symCount_\symRegVal\quad \text{for $\symRegVal \in [1, \symRegRange]$}.
\end{equation}
To get estimators for all other $\symRegVal$, we consider their expectations
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}\right).
\end{equation}
From \eqref{equ:register_value_distribution} we know that $\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)=
e^{-\frac{\symPoissonRate}{\symNumReg}}$ and $\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)=
e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}$, and therefore, we can also write
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg
\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}
\left(1-\symExpectation\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
\end{equation}
and
\begin{equation}
\symExpectation(\symCount'_\symRegVal)
=
\symNumReg
\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}
\left(1-\symExpectation\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}\right),
\end{equation}
which motivates us to use
\begin{equation}
\hat{\symCount}'_\symRegVal
:=
\symNumReg\cdot
\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\cdot
\left(1-\left(\frac{\symCount_0}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
\end{equation}
as estimator for $\symRegVal \leq 0$ and
\begin{equation}
\hat{\symCount}'_\symRegVal
:=
\symNumReg\cdot
\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}
\cdot
\left(1-\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{\symRegRange-\symRegVal}}\right)
\end{equation}
as estimator for $\symRegVal \geq \symRegRange+1$, respectively.

Inserting all these estimators into \eqref{equ:raw_estimate_with_multiplicities} gives
\begin{equation}
\label{equ:correctedestimator}
\symPoissonRateEstimate 
= 
\frac{\symNumReg^2}{2\log(2)\sum_{\symRegVal=-\infty}^\infty \hat{\symCount}'_\symRegVal 2^{-\symRegVal}}
=
\frac{\symNumReg^2}
{2\log(2)
\left(
\symSmallCorrectionFunc(\symCount_0,\symNumReg) + \sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal 2^{-\symRegVal} + \symLargeCorrectionFunc(\symCount_{\symRegRange+1},\symNumReg) 2^{-(\symRegRange+1)}
\right)}
\end{equation}
where $\symSmallCorrectionFunc(\symCount_0)$ and $\symLargeCorrectionFunc(\symCount_{\symRegRange+1})$ replace $\symCount_0$ and $\symCount_{\symRegRange+1}$ in the \emph{raw} estimate \eqref{equ:intermediate_estimate}, respectively. The functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ are given by
\begin{equation}
\symSmallCorrectionFunc(\symCount,\symNumReg) := \symNumReg
\cdot
\left(
\frac{\symCount}{\symNumReg}
+
\sum_{\symRegVal=1}^\infty
\left(\frac{\symCount}{\symNumReg}\right)^{2^\symRegVal} 2^{\symRegVal-1}
\right)
\end{equation}
and
\begin{equation}
\symLargeCorrectionFunc(\symCount,\symNumReg)
:=
\symNumReg
\cdot
\sum_{\symRegVal=1}^\infty
\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{-\symRegVal}}
\cdot
\left(1-\left(1-\frac{\symCount_{\symRegRange+1}}{\symNumReg}\right)^{2^{-\symRegVal}}\right)
\cdot
2^{-(\symRegVal-1)}
\end{equation}

For the special case of linear probabilistic counting with $\symRegRange=0$, we get for the  estimator \eqref{equ:correctedestimator}
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symNumReg}{\log(2)\sum_{\symRegVal=-\infty}^\infty
\left(\frac{\symCount_0}{\symNumReg}\right)^{2^\symRegVal}2^\symRegVal}
=
\frac{\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right)}{\symPowerSeriesFunc\left(\log_2\left(\log\left(\frac{\symNumReg}{\symCount_0}\right)\right)\right)}
\approx
\symNumReg\log\left(\frac{\symNumReg}{\symCount_0}\right).
\end{equation}
Here we used function $\symPowerSeriesFunc$ defined in \eqref{equ:power_series_function} which is known to be approximately equal to 1. Hence, for this special case we get as expected almost the same estimator as given in \eqref{equ:small_range_estimate}.
\subsection{Estimation algorithm}
see \algoref{alg:corrected_raw_estimation}

\begin{algorithm}
\caption{Cardinality estimation algorithm based on the corrected \emph{raw} estimate.}
\label{alg:corrected_raw_estimation}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symCount}$}
\State $\symNumReg \gets \left\Vert\vec{\symCount}\right\Vert_1$
\State $\symZ\gets 0.5\cdot \symLargeCorrectionFunc(\symCount_{\symRegRange+1},\symNumReg)$
\For{$\symRegVal\gets \symRegRange, 1$}
\State $\symZ\gets0.5\cdot\left(\symZ + \symCount_\symRegVal\right)$
\EndFor
\State $\symZ\gets\symZ+\symSmallCorrectionFunc(\symCount_0,\symNumReg)$
\State \Return$\symNumReg^2 / \left(2\log(2)\symZ\right)$
\EndFunction
\\
\Function {$\symSmallCorrectionFunc$}{$\symCount,\symNumReg$}
\If{$\symCount=\symNumReg$}
\State\Return$\infty$
\EndIf
\State $\symX\gets\symCount/\symNumReg$
\State $\symY\gets1$
\State $\symZ\gets\symX$
\Repeat
\State $\symX\gets\symX\cdot\symX$
\State $\symZ'\gets\symZ$
\State $\symZ\gets\symZ + \symX\cdot\symY$
\State $\symY\gets2\cdot\symY$
\Until{$\symZ=\symZ'$}
\State\Return$\symNumReg\cdot\symZ$
\EndFunction
\\
\Function {$\symLargeCorrectionFunc$}{$\symCount,\symNumReg$}
\State $\symX\gets(\symNumReg-\symCount)/\symNumReg$
\State $\symY\gets1$
\State $\symZ\gets0$
\Repeat
\State $\symX\gets\sqrt{\symX}$
\State $\symZ'\gets\symZ$
\State $\symZ\gets\symZ + \left(1-\symX\right)\cdot\symX\cdot\symY$
\State $\symY\gets0.5\cdot\symY$
\Until{$\symZ=\symZ'$}
\State\Return$\symNumReg\cdot\symZ$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\subsection{Performance}
see \autoref{fig:corrected_raw_avg_exec_time}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{corrected_raw_avg_exec_time}
\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively. Both cases, $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ precalculated or calculated on-demand have been considered.}
\label{fig:corrected_raw_avg_exec_time}
\end{figure}

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_avg_exec_time}
\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at 3.3GHz when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
\label{fig:avg_exec_time}
\end{figure}
